---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Kim, Juhwi"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$CLASS,mydata$RINGS)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***(There are two categorical variables to classify sex and class in abalone, the rest of variables including two additionally created Volume and Ratio are continuous in numerical form. Whole, Shuck, and Volume variables may have outliers due to large difference between minimum and maximum values.) ***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
sex_class <- table(mydata$SEX,mydata$CLASS)
addmargins(sex_class)

barplot(sex_class,main="Distribution of Abalone Class by Sex",ylab="Frequency",ylim=c(0,200),xlab="CLASS",beside=TRUE,col=c("Red","Pink","Purple"),legend=rownames(sex_class))

```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: (Infant abalones dominate lower class: A1 and A2 while the highest frequencies of both Male and Female abalone fall in class A3. The ratio between male and female becomes on par in higher class, A4 and A5.)***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
work <- mydata[sample(1:nrow(mydata), 200, replace=FALSE),]
plot(work[, 2:6])


```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
ggplot(data = mydata, aes(x = VOLUME, y = WHOLE)) + 
  geom_point(aes(color = CLASS),size = 1) + 
  ggtitle(bquote('Whole weight vs. Volume')) +
  scale_color_brewer(palette="Set1") +
  ylab("Whole weight (grams)") +
  xlab(bquote('Volume ('~cm^3~')'))
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
ggplot(data = mydata, aes(x = WHOLE, y = SHUCK)) + 
  geom_point(aes(color = CLASS),size = 1,) +
  ggtitle('Shuck weight vs Whole weight') +
  geom_abline(intercept=0, slope=max(mydata$SHUCK/mydata$WHOLE), size=.25, linetype=1) +
  scale_color_brewer(palette="Set1") +
  ylab("Shuck weight (grams)") +
  xlab('Whole weight (grams)') 

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: (Both plots follow similar pattern in that abalones in lower class scatter towards lower left corner indicating the smaller and lower in weight. There are less outliers from lower class in plot (b). Whereas, abalones from higher class seem to be distributed equally and dominates the upper right corner in chart indicating higher volume/weight.)***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
par(mfrow = c(3,3))
hist(mydata$RATIO[mydata$SEX=="I"],main="Infant Ratio",xlab="")
hist(mydata$RATIO[mydata$SEX=="F"],main="Female Ratio",xlab="")
hist(mydata$RATIO[mydata$SEX=="M"],main="Male Ratio",xlab="")

boxplot(mydata$RATIO[mydata$SEX=="I"],main="Infant Ratio")
boxplot(mydata$RATIO[mydata$SEX=="F"],main="Female Ratio")
boxplot(mydata$RATIO[mydata$SEX=="M"],main="Male Ratio")

qqnorm(mydata$RATIO[mydata$SEX=="I"],main="Infant Ratio")
qqline(mydata$RATIO[mydata$SEX=="I"])
qqnorm(mydata$RATIO[mydata$SEX=="F"],main="Female Ratio")
qqline(mydata$RATIO[mydata$SEX=="F"])
qqnorm(mydata$RATIO[mydata$SEX=="M"],main="Male Ratio")
qqline(mydata$RATIO[mydata$SEX=="M"])


```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: (Infant, Female and Male all are exhibit slightly left-skewed distribution, and there are few outlying Ratios that fall above the IQR in all three groups.)***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
infant_out <- mydata[mydata$SEX == "I" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "I"])$out, ]
female_out <- mydata[mydata$SEX == "F" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "F"])$out, ]
male_out <- mydata[mydata$SEX == "M" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "M"])$out, ]
kable(rbind(infant_out, female_out, male_out))

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: (While all Sex categories have numbers of outliers, the highest number of outliers come from the infant classification. Most of outliers, except 1,  are from the A1-A3 classes and only one outlier is classified as A5. While all younger class outliers have ~20% Ratio, the A5 outlier has ~6% Ratio)***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
par(mfrow = c(2,2))
boxplot(mydata$VOLUME ~ mydata$CLASS,ylab="VOLUME")
boxplot(mydata$WHOLE ~ mydata$CLASS,ylab="WHOLE")
plot(mydata$RINGS,mydata$VOLUME,ylab="VOLUME",xlab="RINGS")
plot(mydata$RINGS,mydata$WHOLE,ylab="WHOLE",xlab="RINGS")

```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: (Both volume and whole weight are positively correlated with number of rings as well as class. It appears that higher volume and whole weight are correlated with higher class.  Also, as volume and whole weight increase, number of rings increase as well. so does the number of rings. However, classes A3, A4, and A5 all show very similar IQRs in volume and whole weight, making this correlation not as strong. This indicates that volume and weight are weak predictors of exact class.)***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}

ab_agg <- aggregate(mydata[c('VOLUME', 'SHUCK', 'RATIO')], FUN=mean, by=list(mydata$SEX, mydata$CLASS))
mat_volume <- matrix(ab_agg$VOLUME, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))
mat_shuck <- matrix(ab_agg$SHUCK, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))
mat_ratio <- matrix(ab_agg$RATIO, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))
kable(mat_volume, caption="Volume")
kable(mat_shuck, caption="Shuck")
kable(mat_ratio, caption="Ratio")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
colnames(ab_agg) <- c('Sex', 'Class', 'Volume', 'Shuck', 'Ratio')
ratio_class <- ggplot(data = ab_agg, aes(x = Class, y = Ratio, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Ratio vs. Class") +
  scale_color_brewer(palette="Set1")

volume_class <- ggplot(data = ab_agg, aes(x = Class, y = Volume, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Volume vs. Class") +
  scale_color_brewer(palette="Set1")

shuck_class <- ggplot(data = ab_agg, aes(x = Class, y = Shuck, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Shuck weight vs. Class") +
  scale_color_brewer(palette="Set1")
grid.arrange(ratio_class, volume_class, shuck_class, nrow=3)

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (It appears that Females are larger in volume and shuck weight than Males. What's difference between Female and Male that determines this? Also, it appears that volume and shuck weight peaks at A4 and slightly drops down for A5. What about A5 that drops this variable? Why do infant abalones dominates in Ratio in class A1? How can an infant abalone be ranked A5 in class?)***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
inf_vol <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x=RINGS, y=VOLUME, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="green") + 
  ggtitle("Infant Volume vs. Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 1000)) +
  ylab("Volume") +
  xlab("Rings")
adl_vol <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX == "F")), aes(x=RINGS, y=VOLUME, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="blue") + 
  ggtitle("Adult Volume vs. Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 1000)) +
  ylab("Volume") +
  xlab("Rings")
  
inf_whole <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x=RINGS, y=WHOLE, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="green") + 
  ggtitle("Infant Whole weight vs. Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 400)) +
  ylab("Whole weight") +
  xlab("Rings")
adl_whole <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX == "F")), aes(x=RINGS, y=WHOLE, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="blue") + 
  ggtitle("Adult Whole weight | Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 400)) +
  ylab("Whole weight") +
  xlab("Rings")
grid.arrange(inf_vol, adl_vol, inf_whole, adl_whole, nrow=2)

```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: (The plots suggest that abalone growth follows a left-skewed distribution. The significant portion of the abalone’s development takes place during the later stages. They seem to peak and reach highest weight around 12 rings and then decrease. Adult adults are much more consistent than infant abalones, which is also explained by having more outliers among the Infant abalones. )***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: (Based on the data, it seems to be very difficult to simply distinguish infant vs adult abalones. The data could not support more unique and meaningful relationships in gender, though we found that females weigh more than males or infants. As the abalone ages, the correlating factors became less clear.)***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: (I would fist like clarifications around how sample population was determined, how samples were collected, and where samples come from: one region vs. ramdomly selected across multiple regions.)***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: (Difficulties surrounding observational studies include possible subjectivity and representatives of population. It can also be difficult to derive causality due to external factors that influence the behaviors/results without fully being aware of what are all the factors influencing the data. Causality can't be determined due to researchers being unable to fully controlling input variables, which make causal interpretation difficult without knowing which exact factor influenced the outcome. We can still draw inference using observational studies, especially with the ability to gather larger samples, researchers can capture data from larger sample population. )***
